DNA Methylation Hierarchical Variance Model\nPaper Figures

### 2020-11-24 Paper_Figures.Rmd
Edits to figures

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(gt)
library(rlang)
library(ggpubr)
library(knitr)

@@

stanFile = "FullResultsTCGA_relaxgamma3_postsum3.Rdata"
load(here(stanFile))

prepdir <- "data-stan_rg3ps3"
cat(sprintf("Load prepped data dir: %s (needs to be pre-processed)...",prepdir))
gened_data <- read_csv(here(prepdir, "gened_data.csv"), col_types = list("CHR" = col_character()))
prepped_data <- read_csv(here(prepdir, "prepped_data.csv"))
permed_data <- read_csv(here(prepdir, "permed_data.csv"))

\newpage
\section{Figure 1: Model Overview, Priors, Example Site Fit}
\subsection{Data}
Logit-transformed beta values, from EPIC methylation array pre-processed with minfi Noob algorithm. \newline
\newline Sample counts (total): 48 samples from 21 patients. \newline
By tissue: 6 normal samples, 42 paired bulk tumor samples.
\subsection{Model equation}
Let i=sample, j=patient
$$ y_i = \mu + a_j +  \text{tInd}(i)*(\beta_T + b_j + c_i) $$
$$ a_j \sim N(0,\sigma_P) , b_j \sim N(0,\sigma_{PT}) , c_i \sim N(0,\sigma_T) $$
\newline Model is fit at each individual CpG site using RStan's MCMC sampling algorithm. \newline
Stan Parameters: 4 chains x 2000 iterations (200 warmup) = 7200 total draws. adapt_delta = 0.999

\subsection{Prior Probabilities}
Priors for each parameter (mu, betaT, sigmaP/PT/T) were defined using TCGA 450k array data from colorectal tumor patients with paired multiple tumor and/or normal+tumor samples. \newline Empirical values of mean and standard deviations were assesed for all 450k CpG sites. \newline The resulting distributions were fit with: bimodal gaussian (mu), Cauchy (betaT), or gamma distributions (sigmas). \newline
SigmaE (error variance) prior was defined using a non-informative gamma distribution
\newline Additionally: Prior relaxation was performed for the Cauchy and gamma distribtions by algebraically increasing the variance of the distributions while maintaining the same mode (peak)

@@

# plot mu,betaT,sigmaE,sigmaP,sigmaPT,sigmaT with priors overlaid
pM <- post_prior_histogram(mu_ests, function(x) fBG(x,0.3557,-3.0675, 0.75214,1.1290, 1.39197), function(x) fBG(x,0.3557,-3.0675, sqrt(3)*0.75214,1.1290, sqrt(3)*1.39197), 0.1, c(-5,5), "Mu")
pB <- post_prior_histogram(betaT_ests, function(x) dcauchy(x,-0.04051, 0.19330), function(x) dcauchy(x,-0.04051, 0.5799), 0.1, c(-5,5), "BetaT")
pP <- post_prior_histogram(sigmaP_ests, function(x) dgamma(x, 5.1989, 16.6192), function(x) dgamma(x, 2.7693, 7.0029), 0.05, c(0,2), "SigmaP")
pPT <- post_prior_histogram(sigmaPT_ests, function(x) dgamma(x, 3.5539, 5.5324), function(x) dgamma(x, 2.1457, 2.4819), 0.05, c(0,2), "SigmaPT")
pT <- post_prior_histogram(sigmaT_ests, function(x) dgamma(x, 6.8383, 25.7560), function(x) dgamma(x, 3.3643, 10.430), 0.05, c(0,2), "SigmaT",legFlag=TRUE)
pE <- ggplot() + stat_function(fun=function(x) dgamma(x, 2, 2),geom="line",color="blue") + ggtitle("SigmaE") + xlim(c(0,2)) + theme_classic() + theme(legend.position = "none") + xlab("SigmaE") + ylab("Freq.")

@@

\subsection{CpG Conservation score}
We want to understand which sites of DNA methylation are fundamentally conserved within tumors, i.e. lower variation in methlyation within tumor (sigmaT) relative to the variation within normal healthy tissue (sigmaP). Therefore we choose the log-ratio of sigmaP/sigmaT to represent a conservation score, which is positive when tumor variation is lower than normal variation. We take the median of the posterior distribution for this parameter.

$$\text{score}_\text{med} = \text{Median} (\text{log}\frac{\sigma_{P}}{\sigma_T}) $$

\newpage
\subsection{Example site fit}
![Fit](/Users/kevinmurgas/Documents/Data+ project/Paper Figures/example_fit.png)
\newline \newline
![Posteriors](/Users/kevinmurgas/Documents/Data+ project/Paper Figures/example_posteriors.png)
\newpage
\section{Figure 2: Model Fit Results}
\subsection{Parameter Distributions}
Histograms are median values of posterior distribution
```{r Params_site, echo=FALSE, fig.fullwidth=TRUE, warning=FALSE, comment=NA}
genemask = !(FullAnnotation$UCSC_RefGene_Name=="")
cat(sprintf("N sites: total=%i, gene-associated=%i, non-gene=%i  \n",length(genemask),sum(genemask),length(genemask)-sum(genemask)))

@@

# plot the two conservations score prob sums
pP1 <- post_histogram(prob1_full$median,genemask,0.05,c(-3,3),expression("Median log("~sigma[P]/sigma[T]~")"))
#prob1_proxy = log(sigmaP_full$p50/sigmaT_full$p50)
#pP2 <- post_histogram(prob1_proxy,genemask,0.05,c(-3,3),expression(" log( median" ~sigma[P]/~"median"~sigma[T]~")"))
p_scores <- pP1
print(p_scores)
```

@@

summarize_conservation <- function(data,str) {
  indsin <- str_detect(data$UCSC_RefGene_Group, str) & !is.na(data$UCSC_RefGene_Group)
  indsout <- !str_detect(data$UCSC_RefGene_Group, str) | is.na(data$UCSC_RefGene_Group)
  if (str=="any") {
    indsin <- !is.na(data$UCSC_RefGene_Group)
    indsout <- is.na(data$UCSC_RefGene_Group)
  }
  data.frame(cons_in=mean(data$score_median[indsin], na.rm=TRUE),
         cons_out=mean(data$score_median[indsout],na.rm=TRUE),
         nsites=sum(indsin),
         pval=wilcox.test(data$score_median[indsin],data$score_median[indsout])$p.value)
}

bind_rows({prepped_data %>% summarize_conservation("any") %>% mutate(region="Gene-Associated")},
          {prepped_data %>% summarize_conservation("TSS1500") %>% mutate(region="TSS1500")},
          {prepped_data %>% summarize_conservation("TSS200") %>% mutate(region="TSS200")},
          {prepped_data %>% summarize_conservation("5'UTR|5UTR|5URT") %>% mutate(region="5'UTR")},
          {prepped_data %>% summarize_conservation("1stExon") %>% mutate(region="1stExon")},
          {prepped_data %>% summarize_conservation("Body") %>% mutate(region="Body")},
          {prepped_data %>% summarize_conservation("3'UTR|5UTR") %>% mutate(region="3'UTR")}) %>%
  select(c(region,cons_in,cons_out,nsites,pval)) %>%
  gt() %>%
  fmt_number(columns = vars(cons_in,cons_out,pval), decimals = 3)
```

@@

plotGeneScore <- function(geneStr) {
  CpGscores <- gened_data %>% filter(str_detect(UCSC_RefGene_Name,paste0("^",geneStr,"$"))) %>% pull(score_median)

  ggplot(prepped_data) + geom_histogram(aes(x=score_median, y=..count../nrow(prepped_data)), binwidth=0.05) + xlim(-3,3) + ggtitle(sprintf("%s: CpG scores (%i sites)",geneStr, length(CpGscores))) + xlab("CpG conservation score") + ylab("Freq.") + theme_classic() + theme(legend.position = "none") + geom_vline(xintercept=CpGscores, color="red", alpha=0.5, size=0.2) + geom_vline(xintercept=mean(CpGscores), color="blue", alpha=0.8, size=2)
}

cat("\n")
plotGeneScore("TP53")
cat("\n")
plotGeneScore("TTN")
cat("\n")
plotGeneScore("HLA-A")
```

@@

cancergenes <- read_csv("/Users/kevinmurgas/Documents/Data+ project/Cancer gene lists/SangerCensus_COSMIC_CRC.csv")
genescores <- gene_summary %>% filter(UCSC_RefGene_Name %in% (cancergenes %>% pull(`Gene Symbol`))) %>% pull(score_median)
ggplot(gene_summary) + geom_histogram(aes(x=score_median, y=..count../nrow(gene_summary)), binwidth=0.05) + xlim(-3,3) + ggtitle(sprintf("Gene Scores of %i CRC Genes", length(genescores))) + xlab("Gene conservation score") + ylab("Freq.") + theme_classic() + theme(legend.position = "none") + geom_vline(xintercept=genescores, color="blue", alpha=0.5, size=0.2)
cat("\n")

# source 2
cancergenes <- read_csv("/Users/kevinmurgas/Documents/Data+ project/Cancer gene lists/CRCgenes_atlasgeneticsoncology.csv")
genescores <- gene_summary %>% filter(UCSC_RefGene_Name %in% (cancergenes %>% pull(CancerGenes))) %>% pull(score_median)
ggplot(gene_summary) + geom_histogram(aes(x=score_median, y=..count../nrow(gene_summary)), binwidth=0.05) + xlim(-3,3) + ggtitle(sprintf("Gene Scores of %i CRC Genes", length(genescores))) + xlab("Gene conservation score") + ylab("Freq.") + theme_classic() + theme(legend.position = "none") + geom_vline(xintercept=genescores, color="blue", alpha=0.1, size=0.2)
```

@@

enrich_sigmaP <- permed_data %>%
  #filter(sigmaP_v3 < quantile(sigmaP_v3, 0.05, na.rm = TRUE)) %>%
  filter(sigmaP_v3 < 0.05) %>%
  pull(UCSC_RefGene_Name) %>% tidy_gsea()
enrich_sigmaPT <- permed_data %>%
  #filter(sigmaPT_v3 < quantile(sigmaPT_v3, 0.05, na.rm = TRUE)) %>%
  filter(sigmaPT_v3 < 0.05) %>%
  pull(UCSC_RefGene_Name) %>% tidy_gsea()
enrich_sigmaT <- permed_data %>%
  #filter(sigmaT_v3 < quantile(sigmaT_v3, 0.05, na.rm = TRUE)) %>%
  filter(sigmaT_v3 < 0.05) %>%
  pull(UCSC_RefGene_Name) %>% tidy_gsea()
#enrich_score_prob <- permed_data %>%
#  filter(score_prob_v3 > quantile(score_prob_v3, 0.95, na.rm = TRUE)) %>%
#  pull(UCSC_RefGene_Name) %>% tidy_gsea()
#enrich_score_mean <- permed_data %>%
#  filter(score_mean_v3 > quantile(score_mean_v3, 0.95, na.rm = TRUE)) %>%
#  pull(UCSC_RefGene_Name) %>% tidy_gsea()
enrich_score_median <- permed_data %>%
  #filter(score_median_v3 > quantile(score_median_v3, 0.95, na.rm = TRUE)) %>%
  filter(score_median_v3 > 0.95) %>%
  pull(UCSC_RefGene_Name) %>% tidy_gsea()
enrich_score_median2 <- permed_data %>%
  #filter(score_median2_v3 > quantile(score_median2_v3, 0.95, na.rm = TRUE)) %>%
  filter(score_median2_v3 > 0.95) %>%
  pull(UCSC_RefGene_Name) %>% tidy_gsea()
enrich_score_reg <- permed_data %>%
  #filter(score_reg_v3 > quantile(score_reg_v3, 0.95, na.rm = TRUE)) %>%
  filter(score_reg_v3 > 0.95) %>%
  pull(UCSC_RefGene_Name) %>% tidy_gsea()
```

@@

cat("Summarize by gene, taking mean value of all sites in gene")
ptm<-proc.time()
gene_summary <- gened_data %>%
  filter(!is.na(UCSC_RefGene_Name)) %>%
  gene_summarize(UCSC_RefGene_Name, c(mu:sigmaT,score_median:score_reg))
cat(proc.time()-ptm)

@@

# plot: n_sites, mu, betaT, sigmaP, sigmaPT, sigmaT
n_cutlo = 5
n_cuthi = quantile(gene_summary$n, 0.99)
pN <- post_histogram(gene_summary$n,1,c(0,200),paste0("N sites (max = ",max(gene_summary$n),")")) + geom_vline(xintercept=n_cutlo-0.5,linetype = "dashed",color="red") + annotate(geom="text",x=80,y=0.03,size=2,label=paste0("# genes <",n_cutlo," sites:\n", sum(gene_summary$n<n_cutlo)," / ",nrow(gene_summary))) +
geom_vline(xintercept=n_cuthi,linetype = "dashed",color="red") +annotate(geom="text",x=120,y=0.013,size=2,label=sprintf("# genes >%0.1f sites:\n %i / %i (99th pctile)", n_cuthi,sum(gene_summary$n>n_cuthi),nrow(gene_summary)))
gene_summary <- gene_summary %>% filter(n>=n_cutlo) %>% filter(n<n_cuthi)
pM <- post_histogram(gene_summary$mu,0.1,c(-5,5),"Mu")
pB <- post_histogram(gene_summary$betaT,0.1,c(-5,5),"BetaT")
pP <- post_histogram(gene_summary$sigmaP,0.05,c(0,2),"SigmaP")
pPT <- post_histogram(gene_summary$sigmaPT,0.05,c(0,2),"SigmaPT")
pT <- post_histogram(gene_summary$sigmaT,0.05,c(0,2),"SigmaT")
pP1 <- post_histogram(gene_summary$score_median,0.05,c(-3,3),"Score_median")
pP2 <- post_histogram(gene_summary$score_median2,0.05,c(-3,3),"Score_median2")
pP3 <- post_histogram(gene_summary$score_reg,0.05,c(-3,3),"Score_reg")

#pP1 <- post_histogram(gene_summary$score_prob,0.01,c(0,1),"Score_prob")
#pP2 <- post_histogram(gene_summary$score_mean,0.05,c(-3,3),"Score_mean")

# arrange each parameter as a subplot
p_all <- ggarrange(pN,pM,pB,pP,pPT,pT,pP1,pP2,pP3,hjust=0,nrow=3,ncol=3)
print(p_all)
rm(pN,pM,pB,pP,pPT,pT,pP1,pP2,pP3)

@@

cat("Number of enrichment pathways (probs):")
c(score_median = length(enrich_score_median$Description),
  score_median2 = length(enrich_score_median2$Description),
  score_reg = length(enrich_score_reg$Description)) %>%
  sort()

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggpubr)

library(gt)
library(rlang)
library(knitr)
library(gtools)

@@

stanFile = "FullResultsTCGA_relaxgamma3.Rdata"
load(here(stanFile))

prepdatafolder = "data_final_rg3"
cat(sprintf("Load prepped data dir: %s (needs to be pre-processed)...",prepdir))
gened_data <- read_csv(here(prepdatafolder, "gened_data.csv"), col_types = list("CHR" = col_character()))
prepped_data <- read_csv(here(prepdatafolder, "prepped_data.csv"))
permed_data <- read_csv(here(prepdatafolder, "permed_data.csv"))

# # plot beta/M values
# all_beta = FullAnnotation[,-c(1,2,3,4,5,6,7,8,9)]
# mean_beta = rowMeans(all_beta)
# all_M = logit(all_beta)
# mean_M = rowMeans(all_M)
# p_beta <- data.frame(value=mean_beta) %>% gather %>% ggplot() + geom_histogram(aes(x=value),binwidth=0.01,fill="lightgray",color="grey50") + theme_classic() + xlim(c(0,1)) + theme(aspect.ratio = 1) + xlab("Beta value")
# p_M <- data.frame(value=mean_M) %>% gather %>% ggplot() + geom_histogram(aes(x=value),binwidth=0.1,fill="lightgray",color="grey50") + theme_classic() + xlim(c(-5,5)) + theme(aspect.ratio = 1) + xlab("M value")
# p_BM <- ggarrange(p_beta, p_M, nrow = 1, ncol = 2)
# # ggsave("betaMdists.svg",p_BM)
```

\newpage
\section{Model Overview, Priors, Example Site Fit}

\subsection{Data}
Logit-transformed beta values, from EPIC methylation array pre-processed with minfi Noob algorithm. \newline
\newline Sample counts (total): 48 samples from 21 patients. \newline
By tissue: 6 normal samples, 42 paired bulk tumor samples.

\subsection{Model equation}
Let i=patient, j=sample (0=normal, 1,2=tumor sample 1,2)
$$ y_{ij} = \mu + \alpha_i +  (1-\delta_{j0})*(\nu + \beta_i + \gamma_{ij}) + \epsilon_{ij} $$
$$ \alpha_i \sim N(0,\sigma_P) , \beta_i \sim N(0,\sigma_{PT}) , \gamma_{ij} \sim N(0,\sigma_T) $$
\newline Model is fit at each individual CpG site using RStan's MCMC sampling algorithm. \newline
Stan Parameters: 4 chains x 2000 iterations (200 warmup) = 7200 total draws. adapt_delta = 0.999

\subsection{Prior Probabilities}
Priors for each parameter (mu, nu, sigmaP/PT/T) were defined using TCGA 450k array data from colorectal tumor patients with paired multiple tumor and/or normal+tumor samples. \newline Empirical values of mean and standard deviations were assesed for all 450k CpG sites. \newline The resulting distributions were fit with: bimodal gaussian (mu), Cauchy (betaT), or gamma distributions (sigmas). \newline
SigmaE (error variance) prior was defined using a non-informative gamma distribution: gamma(2,2).
\newline Additionally: Prior relaxation was performed by algebraically increasing the variance of the distributions (or each Gaussian component in the case of mu) while maintaining the same mode (peak).

@@

# plot mu,betaT,sigmaE,sigmaP,sigmaPT,sigmaT with priors overlaid
pM <- post_prior_histogram(mu_ests, function(x) fBG(x,0.3557,-3.0675, 0.75214,1.1290, 1.39197), function(x) fBG(x,0.3557,-3.0675, sqrt(3)*0.75214,1.1290, sqrt(3)*1.39197), 0.1, c(-5,5), "Mu")
pB <- post_prior_histogram(betaT_ests, function(x) dcauchy(x,-0.04051, 0.19330), function(x) dcauchy(x,-0.04051, 0.5799), 0.1, c(-5,5), "Nu")
pP <- post_prior_histogram(sigmaP_ests, function(x) dgamma(x, 5.1989, 16.6192), function(x) dgamma(x, 2.7693, 7.0029), 0.05, c(0,2), "SigmaP")
pPT <- post_prior_histogram(sigmaPT_ests, function(x) dgamma(x, 3.5539, 5.5324), function(x) dgamma(x, 2.1457, 2.4819), 0.05, c(0,2), "SigmaPT")
pT <- post_prior_histogram(sigmaT_ests, function(x) dgamma(x, 6.8383, 25.7560), function(x) dgamma(x, 3.3643, 10.430), 0.05, c(0,2), "SigmaT",legFlag=TRUE)
pE <- ggplot() + stat_function(fun=function(x) dgamma(x, 2, 2),geom="line",color="blue") + ggtitle("SigmaE") + xlim(c(0,2)) + theme_classic() + theme(legend.position = "none") + xlab("SigmaE") + ylab("Freq.")

@@

\subsection{CpG Conservation score}
We want to understand which sites of DNA methylation are fundamentally conserved within tumors, i.e. lower variation in methlyation within tumor (sigmaT) relative to the variation within normal healthy tissue (sigmaP). Therefore we choose the log-ratio (base 2) of sigmaP/sigmaT to represent a relative conservation score, which is positive when tumor variation is lower than normal variation. We take the median of the posterior distribution constructed from the two sigmaP,T posteriors for this score.

$$\text{score}_\text{med} = \text{Median} (\text{log}_2\frac{\sigma_{P}}{\sigma_T}) $$

\newpage
\subsection{Figure 2A: Example site fit, 2B: Posteriors}
![Fit](/Users/kevinmurgas/Documents/Data+ project/Manuscript/example_fit.png)
\newline \newline
![Posteriors](/Users/kevinmurgas/Documents/Data+ project/Manuscript/example_posteriors.png)
\newpage
\section{Figure 2: Model Fit Results}
\subsection{Parameter Distributions}
Histograms are median values of posterior distribution
```{r Params_site, echo=FALSE, fig.fullwidth=TRUE, warning=FALSE, comment=NA}
genemask = !(FullAnnotation$UCSC_RefGene_Name=="")
cat(sprintf("N sites: total=%i, gene-associated=%i, non-gene=%i  \n",length(genemask),sum(genemask),length(genemask)-sum(genemask)))

@@

# overlaid transparent histograms
post_histogram <- function(postvals,mask,bw,titlestr) {
  # stacked histogram colored by gene-associated sites
  # and prior curve overlaid
  plot_df <- data.frame(val=postvals, gene_assoc=mask)
  gene_assoc_df <- plot_df %>% filter(gene_assoc)
  not_gene_assoc_df <- plot_df %>% filter(!gene_assoc)
  p1 <- ggplot(plot_df, aes(x=val)) +
    geom_histogram(data=not_gene_assoc_df, aes(y=..density..,fill="#00BFC4",alpha=0.4), binwidth=bw) +
    geom_histogram(data=gene_assoc_df, aes(y=..density..,fill="#F8766D",alpha=0.4), binwidth=bw) +
    ggtitle(titlestr) + xlab(titlestr) + ylab("Density") + theme_classic() 
  p1
}

# plot the two conservations score prob sums
genemask = !(FullAnnotation$UCSC_RefGene_Name=="")
p_scores <- post_histogram(CpGscore_full$median,genemask,0.05,"CpG Conservation score") +
  theme(aspect.ratio=1/4,legend.position = "none") +
  scale_x_continuous(limits=c(-3,3),breaks=c(-3,-2,-1,0,1,2,3))
print(p_scores)
# ggsave("CpGscores_overlay.svg")

# plot gene scores and adjusted bootstrap distribution
p1 <- permed_data %>% ggplot(aes(x=score_median)) + geom_histogram(aes(y=..count..), binwidth=0.05, fill="lightgray",color = "grey50") + xlim(c(-1,2.5)) + theme_classic() + ylab("Number of genes") + theme(aspect.ratio = 1)
p2 <- permed_data %>% ggplot(aes(x=score_median_v3)) + geom_histogram(aes(y=..count..), binwidth=0.02, fill="lightgray", color="grey50") + xlim(c(0,1)) + theme_classic() + ylab("Number of genes") + theme(aspect.ratio = 1)
p_genes <- ggarrange(p1,p2,nrows=2,ncol=1)
# ggsave("gene_scores_bootstrap.svg")
```

@@

# stats for cpg island relations
# summarize mean of each group, do kruskal wallis and post hoc Tukey
my_data = prepped_data %>% select(score_median, Relation_to_UCSC_CpG_Island) %>%
  group_by(ir = Relation_to_UCSC_CpG_Island) %>%
  mutate(ir = case_when(is.na(ir) ~ "Sea",
                        str_detect(ir, "^S") ~ str_replace(ir, "^S_", ""),
                        str_detect(ir, "^N") ~ str_replace(ir, "^N_", ""),
                        TRUE ~ ir))
my_data %>%
  summarise_at(vars(score_median), mean, na.rm = TRUE)
kruskal.test(score_median ~ ir, data = my_data)
#install.packages("MultNonParam")
library(MultNonParam)
tukey.kruskal.test(my_data$score_median, my_data$ir, alpha = 0.001)

# stats for functional regions
summarize_conservation <- function(data,str) {
  indsin <- str_detect(data$UCSC_RefGene_Group, str) & !is.na(data$UCSC_RefGene_Group)
  indsout <- !str_detect(data$UCSC_RefGene_Group, str) | is.na(data$UCSC_RefGene_Group)
  if (str=="any") {
    indsin <- !is.na(data$UCSC_RefGene_Group)
    indsout <- is.na(data$UCSC_RefGene_Group)
  }
  data.frame(cons_in=mean(data$score_median[indsin], na.rm=TRUE),
         cons_out=mean(data$score_median[indsout],na.rm=TRUE),
         nsites=sum(indsin),
         pval=wilcox.test(data$score_median[indsin],data$score_median[indsout])$p.value)
}

bind_rows({prepped_data %>% summarize_conservation("any") %>% mutate(region="Gene-Associated")},
          {prepped_data %>% summarize_conservation("TSS1500") %>% mutate(region="TSS1500")},
          {prepped_data %>% summarize_conservation("TSS200") %>% mutate(region="TSS200")},
          {prepped_data %>% summarize_conservation("5'UTR|5UTR|5URT") %>% mutate(region="5'UTR")},
          {prepped_data %>% summarize_conservation("1stExon") %>% mutate(region="1stExon")},
          {prepped_data %>% summarize_conservation("Body") %>% mutate(region="Body")},
          {prepped_data %>% summarize_conservation("3'UTR|3UTR") %>% mutate(region="3'UTR")}) %>%
  select(c(region,cons_in,cons_out,nsites,pval)) %>%
  gt() %>%
  fmt_number(columns = vars(cons_in,cons_out,pval), decimals = 3)

# new stats
FRdata = data.frame(CpGscore = log2(prepped_data$sigmaP / prepped_data$sigmaT) ) %>%
  mutate(FR = pull(prepped_data,UCSC_RefGene_Group)) %>%
  mutate(FR=if_else(is.na(FR),"NA",FR)) %>%
  mutate(iTSS200 = str_detect(FR,"TSS200"),
         i5UTR = str_detect(FR,"5'UTR|5UTR|5URT"),
         iBody = str_detect(FR,"Body"),
         i3UTR = str_detect(FR,"3'UTR|3UTR"),
         iTSS1500 = str_detect(FR,"TSS1500"),
         i1stEx = str_detect(FR,"1stExon"),
         iExBnd = str_detect(FR,"ExonBnd"))
fr.lm = lm(formula = CpGscore ~ iTSS200 + i5UTR + iBody + i3UTR + iTSS1500 + i1stEx + iExBnd, data = FRdata)
#fr.lm = lm(formula = CpGscore ~ iTSS200 + i5UTR + iBody + i3UTR, data = FRdata)
summary(fr.lm)

library(rcompanion)
fr.srh = scheirerRayHare(formula = CpGscore ~ iTSS200 + i5UTR + iBody + i3UTR + iTSS1500 + i1stEx + iExBnd, data = FRdata)
```

@@

plotGeneScore <- function(geneStr,norm) {
  CpGscores <- gened_data %>% filter(str_detect(UCSC_RefGene_Name,paste0("^",geneStr,"$"))) %>% pull(score_median)

  ggplot(prepped_data) + geom_histogram(aes(x=score_median, y=..count../nrow(prepped_data)), binwidth=0.05,fill="lightgray") +
    xlim(-3,3) + ggtitle(sprintf("%s: CpG scores (%i sites)",geneStr, length(CpGscores))) +
    xlab("CpG conservation score") + ylab("Freq.") + theme_classic() + theme(legend.position = "none") +
    #geom_vline(xintercept=CpGscores, color="red", alpha=0.5, size=0.2) +
    geom_histogram(data=data.frame(score=CpGscores),aes(x=score,y=norm*..count../length(CpGscores)),binwidth=0.05, fill="blue",alpha=0.5) +
    geom_vline(xintercept=mean(CpGscores), color="blue",linetype="dashed", size=1)
}

cat("\n")
#plotGeneScore("TP53",0.5)
plotGeneScore("TUBA1A",0.285)
#ggsave("genescore_TUBA1A.svg")
cat("\n")
plotGeneScore("TTN",0.37)
#ggsave("genescore_TTN.svg")
cat("\n")
plotGeneScore("HLA-A",0.26)
#ggsave("genescore_HLA-A.svg")

# stats for each gene
# do one sample wilcoxon rank sum
gened_data %>% filter(str_detect(UCSC_RefGene_Name,paste0("^TUBA1A$"))) %>% pull(score_median) %>% wilcox.test(mu = 0, alternative = "two.sided")
gened_data %>% filter(str_detect(UCSC_RefGene_Name,paste0("^TTN$"))) %>% pull(score_median) %>% wilcox.test(mu = 0, alternative = "two.sided")
gened_data %>% filter(str_detect(UCSC_RefGene_Name,paste0("^HLA-A$"))) %>% pull(score_median) %>% wilcox.test(mu = 0, alternative = "two.sided")
```

@@

cancergenes1 <- read_csv("/Users/kevinmurgas/Documents/Data+ project/Cancer gene lists/SangerCensus_COSMIC_CRC.csv")
genescores <- gene_summary %>% filter(UCSC_RefGene_Name %in% (cancergenes1 %>% pull(`Gene Symbol`))) %>% pull(score_median)
ggplot(gene_summary) + geom_histogram(aes(x=score_median, y=..count../nrow(gene_summary)), binwidth=0.05) + xlim(-3,3) + ggtitle(sprintf("Gene Scores of %i CRC Genes", length(genescores))) + xlab("Gene conservation score") + ylab("Freq.") + theme_classic() + theme(legend.position = "none") + geom_vline(xintercept=genescores, color="blue", alpha=0.5, size=0.2)
cat("\n")

# source 2
cancergenes2 <- read_csv("/Users/kevinmurgas/Documents/Data+ project/Cancer gene lists/CRCgenes_atlasgeneticsoncology.csv")
genescores <- gene_summary %>% filter(UCSC_RefGene_Name %in% (cancergenes2 %>% pull(CancerGenes))) %>% pull(score_median)
ggplot(gene_summary) + geom_histogram(aes(x=score_median, y=..count../nrow(gene_summary)), binwidth=0.05) + xlim(-3,3) + ggtitle(sprintf("Gene Scores of %i CRC Genes", length(genescores))) + xlab("Gene conservation score") + ylab("Freq.") + theme_classic() + theme(legend.position = "none") + geom_vline(xintercept=genescores, color="blue", alpha=0.1, size=0.2)

# compute % overlap of database genes and significant genes (By adj bootstrap)
n1 = sum(permed_data %>% filter(score_median_v3>0.95) %>% pull(UCSC_RefGene_Name) %in% (cancergenes1 %>% pull(`Gene Symbol`)))
n2 = sum(permed_data %>% filter(score_median_v3>0.95) %>% pull(UCSC_RefGene_Name) %in% (cancergenes2 %>% pull(CancerGenes)))

# plot violins of: all genes, COSMIC, AGO
perm_temp <- permed_data %>% select(c(UCSC_RefGene_Name, score_median,score_median_v3,n))
temp <- rbind(perm_temp %>% mutate(group="all"),
              perm_temp %>% filter(UCSC_RefGene_Name %in% (cancergenes1 %>% pull(`Gene Symbol`))) %>% mutate(group="COSMIC"),
              perm_temp %>% filter(UCSC_RefGene_Name %in% (cancergenes2 %>% pull(CancerGenes))) %>% mutate(group="AGO")
              )
temp %>% mutate(group = factor(group, levels = c("COSMIC", "AGO", "all"))) %>%
  ggplot() + geom_violin(aes(y=score_median,x=group,fill=group), draw_quantiles = 0.5) + scale_y_continuous(limits=c(-1,2),breaks = c(-1,0,1,2)) + theme_classic() + xlab("Database") + ylab("Gene Conservation Score")
# ggsave("database_violins.svg")

# try with histogram
plotDBScore <- function(geneStr,norm) {
  CpGscores <- gened_data %>% filter(str_detect(UCSC_RefGene_Name,paste0("^",geneStr,"$"))) %>% pull(score_median)

  ggplot(prepped_data) + geom_histogram(aes(x=score_median, y=..count../nrow(prepped_data)), binwidth=0.05,fill="lightgray") +
    xlim(-3,3) + ggtitle(sprintf("%s: CpG scores (%i sites)",geneStr, length(CpGscores))) +
    xlab("CpG conservation score") + ylab("Freq.") + theme_classic() + theme(legend.position = "none") +
    #geom_vline(xintercept=CpGscores, color="red", alpha=0.5, size=0.2) +
    geom_histogram(data=data.frame(score=CpGscores),aes(x=score,y=norm*..count../length(CpGscores)),binwidth=0.05, fill="blue",alpha=0.5) +
    geom_vline(xintercept=mean(CpGscores), color="blue",linetype="dashed", size=1)
}

@@

enrich_sigmaP <- permed_data %>%
  #filter(sigmaP_v3 < quantile(sigmaP_v3, 0.05, na.rm = TRUE)) %>%
  filter(sigmaP_v3 < 0.05) %>%
  pull(UCSC_RefGene_Name) %>% tidy_gsea()
enrich_sigmaPT <- permed_data %>%
  #filter(sigmaPT_v3 < quantile(sigmaPT_v3, 0.05, na.rm = TRUE)) %>%
  filter(sigmaPT_v3 < 0.05) %>%
  pull(UCSC_RefGene_Name) %>% tidy_gsea()
enrich_sigmaT <- permed_data %>%
  #filter(sigmaT_v3 < quantile(sigmaT_v3, 0.05, na.rm = TRUE)) %>%
  filter(sigmaT_v3 < 0.05) %>%
  pull(UCSC_RefGene_Name) %>% tidy_gsea()
enrich_score_median <- permed_data %>%
  #filter(score_median_v3 > quantile(score_median_v3, 0.95, na.rm = TRUE)) %>%
  filter(score_median_v3 > 0.95) %>%
  pull(UCSC_RefGene_Name) %>% tidy_gsea()
```

@@

cat("Summarize by gene, taking mean value of all sites in gene")
ptm<-proc.time()
gene_summary <- gened_data %>%
  filter(!is.na(UCSC_RefGene_Name)) %>%
  gene_summarize(UCSC_RefGene_Name, c(mu:sigmaT,score_median))
cat(proc.time()-ptm)

@@

# plot: n_sites, mu, betaT, sigmaP, sigmaPT, sigmaT
n_cutlo = 5
n_cuthi = quantile(gene_summary$n, 0.99)
pN <- post_histogram(gene_summary$n,1,c(0,200),paste0("N sites (max = ",max(gene_summary$n),")")) + geom_vline(xintercept=n_cutlo-0.5,linetype = "dashed",color="red") + annotate(geom="text",x=80,y=0.03,size=2,label=paste0("# genes <",n_cutlo," sites:\n", sum(gene_summary$n<n_cutlo)," / ",nrow(gene_summary))) +
geom_vline(xintercept=n_cuthi,linetype = "dashed",color="red") +annotate(geom="text",x=120,y=0.013,size=2,label=sprintf("# genes >%0.1f sites:\n %i / %i (99th pctile)", n_cuthi,sum(gene_summary$n>n_cuthi),nrow(gene_summary)))
gene_summary <- gene_summary %>% filter(n>=n_cutlo) %>% filter(n<n_cuthi)
pM <- post_histogram(gene_summary$mu,0.1,c(-5,5),"Mu")
pB <- post_histogram(gene_summary$betaT,0.1,c(-5,5),"BetaT")
pP <- post_histogram(gene_summary$sigmaP,0.05,c(0,2),"SigmaP")
pPT <- post_histogram(gene_summary$sigmaPT,0.05,c(0,2),"SigmaPT")
pT <- post_histogram(gene_summary$sigmaT,0.05,c(0,2),"SigmaT")
pP1 <- post_histogram(gene_summary$score_median,0.05,c(-3,3),"Score_median")
# arrange each parameter as a subplot
p_all <- ggarrange(pN,pM,pB,pP,pPT,pT,pP1,hjust=0,nrow=3,ncol=3)
print(p_all)
rm(pN,pM,pB,pP,pPT,pT,pP1)

@@

cat("Number of enrichment pathways (probs):")
c(score_median = length(enrich_score_median$Description)) %>%
  sort()

list(
  score_median = enrich_score_median$Description
) %>%
  UpSetR::fromList() %>%
  UpSetR::upset(sets.bar.color = RColorBrewer::brewer.pal(3, "Set2")[c( 1)])
grid::grid.text("Figure 3B: Frequency of overlap between pathways that are called \nas most conserved for each conservation score.", x=0.65, y=0.05, gp=grid::gpar(fontsize=10))
```
